Parsing Postfix log files

### 2007-07-22 doc/logparser.tex
Updated coverage section now that there are no unknown mails in the state table. Documented logging HELOs.

XXX MAYBE A PRIOR ART SECTION\@?

@@

Coverage of mails is much more difficult to determine accurately than
coverage of lines.  The parser can dump its state tables in a human
readable form; examining these tables with reference to the log files is
the best way to detect mails which were not handled properly (all the
complications discussed in section~\ref{additional complications} were
diagnosed and sometimes detected in this way).  The parser issues warnings
when it detects any errors, some of which may alert the user to a problem,
e.g.\ when a queueid is reused before the previous mail is fully dealt
with, when a queueid or pid is not found,\footnote{There will often be a
few warnings about missing queueids or pids in the first few hundred or
thousand log lines because the earlier log lines are in the previous log
file; loading the saved state from the previous log file will solve this
problem} or when there are problems tracking a child mail (see
section~\ref{tracking re-injected mail}).  There should be few or no
warnings, and the state table should only contain entries for mails which
had yet to be delivered when the logfiles ended.

At the time of writing the parser is being tested with 93 log files.  There
are 5 warnings produced, but because the parser errs on the side of
producing more warnings rather than fewer those 5 warnings represent only 3
instances of 1 problem: 3 connections started before the first log file, so
their initial log entries are missing, leading to warnings when their log
lines are parsed.

The state tables contain entries for mails not yet delivered when the
parser finishes execution.  Ideally all they should contain are mails which
are awaiting delivery after the period covered by the logs, though they may
also contain mails whose initial entries are not contained in the logs.
Any other entries are evidence of a failure in parsing or an aberration in
the logs.  After parsing the 93 test log files the state tables contain 22
entries, breaking down into:

@@

    \item 2 connections which started only seconds before the logs ended
        and had not completed.

@@

    \item 4 mails present for unknown reasons.

@@

@@

    \item It appears that the hostname used in the HELO command is not
        logged if the mail is accepted.\footnote{Tested with Postfix 2.2.10
        and 2.3.11; this may possibly have changed in Postfix 2.4.}  It
        should be reasonably simple to write a policy server which causes
        Postfix to log a warning containing the HELO hostname when the DATA
        command is accepted.

>>>>

@@

Coverage of mails is much more difficult to determine accurately than
coverage of lines.  The parser can dump its state tables in a human
readable form; examining these tables with reference to the log files is
the best way to detect mails which were not handled properly (many of the
complications discussed in section~\ref{additional complications} were
detected in this way).  The parser issues warnings when it detects any
errors, some of which may alert the user to a problem, e.g.\ when a queueid
is reused before the previous mail is fully dealt with, when a queueid or
pid is not found,\footnote{There will often be warnings about missing
queueids or pids in the first few hundred or thousand log lines because the
earlier log lines are in the previous log file; loading the saved state
from the previous log file will solve this problem.} or when there are
problems tracking a child mail (see section~\ref{tracking re-injected
mail}).  There should be few or no warnings, and the state table should
only contain entries for mails which had yet to be delivered when the
logfiles ended.

At the time of writing the parser is being tested with 93 log files.  There
are 5 warnings produced, but because the parser errs on the side of
producing more warnings rather than fewer those 5 warnings represent 3
instances of 1 problem: 3 connections started before the first log file, so
their initial log entries are missing, leading to warnings when their log
lines are parsed.

The state tables contain entries for mails not yet delivered when the
parser finishes execution.  Ideally all they should contain are mails which
are awaiting delivery after the period covered by the logs, though they may
also contain mails whose initial entries are not contained in the logs.
Any other entries are evidence of a failure in parsing or an aberration in
the logs.  After parsing the 93 test log files the state tables contain 18
entries, breaking down into:

@@

    \item 1 connection which started only seconds before the logs ended and
        had not yet completed.

    \item 1 mail which had been accepted only seconds before the logs ended
        and had not yet been delivered.

@@

@@

The XXX connections and mails handled by Postfix during this period are
handled correctly by the parser.

@@

    \item It appears that the hostname used in the HELO command is not
        logged if the mail is accepted.\footnote{Tested with Postfix 2.2.10
        and 2.3.11; this may possibly have changed in Postfix 2.4.}  It is
        quite easy to rectify this:

            \item Create \texttt{/etc/postfix/log\_helo.pcre}
                containing:\newline
                \hspace*{2em}\texttt{/./~~~~WARN~Logging~HELO}

            \item Modify \texttt{smtpd\_sender\_restrictions} in
                \texttt{/etc/postfix/main.cf} to contain\newline
                \hspace*{2em}\texttt{check\_client\_access~/etc/postfix/log\_helo.pcre}

        The purpose of logging the HELO in
        \texttt{smtpd\_sender\_restrictions} is to only log once per mail.
